---
#WeGo TSP Project 
#
---
**Main question: Does TSP reduce the probability that a bus leaves a stop late?** 

FILTER CRITERIA: 1) TIME/DATE, 
2) Route portion (CH46, 28&CHARL, MCC5_1) --> additional criteria, see route section 

**TSP condition column: on/off (0/1)**
TSP off= Anything before 02-03-2025 is TSP OFF
TSP on= 2025-02-03 12pm to 2025-02-10 12pm 
TSP ON [schedlue Cond]= 2025-02-10 12 pm to 2025-04-28
** TSP Schedule-Conditional Priority Begins (Only buses more than 2 minutes late receive priority)
TSP OFF= 2025-04-28 12pm to 2025-05-05 12pm 
TSP ON= 2025-05-05 12pm to 2025-05-12 at 12pm
TSP on [headway cond]= 2025-05-12 12pm to CURRENT/LAST DATE
**TSP Headway-Conditional TSP Priority Begins (Only gapped buses with actual leading headway more than 120% of scheduled headway receive priority)

Strategy: use the case_when function 
https://www.rdocumentation.org/packages/dplyr/versions/1.0.10/topics/case_when

lubridate: helpful for date-and-time conversions (https://rstudio.github.io/cheatsheets/html/lubridate.html)
  
```{r}
#load in libraries 
library(tidyverse)
library(lubridate)

#load in data
wego <- read_csv("../data/Route 50 Timepoint and Headway Data, 1-1-2023 through 5-12-2025.csv")

#view data
wego|> head()

```

Additional route criteria: 
* Only portions of the route without any TSP is WALM, HLWD
* WHBG route only has TSP if it is going 'to downtown' 
* otherwise, all other route stops are TSP-eligible 
```{r}

#Create a tsp-eligible column to identify the routes
wego <- wego |> mutate(tsp_eligible = case_when(
  #Flag 1 for all rows except the three routes not included 
  TIME_POINT_ABBR != 'WALM' & TIME_POINT_ABBR != 'HLWD' & TIME_POINT_ABBR != 'WHBG'~ 1,
  #flag 1 for WHBG only if it is going in the correct direction
  TIME_POINT_ABBR == 'WHBG' & ROUTE_DIRECTION_NAME == 'TO DOWNTOWN' ~1, 
  #All other routes (aka ones that are excluded from tsp) marked as a 0
  .default = 0), 
    #place the identifier column at the front of the dataframe for easier reference 
  .before = CALENDAR_ID)

#view updated results
wego |> head()
```
```{r}
#Create a timestamp column that combines the date and scheduled time, then make it a datetime object
wego <- wego |> 
  mutate(timestamp = paste(DATE, SCHEDULED_TIME), 
                       .before = DATE) |>
  mutate(timestamp = ymd_hms(timestamp))

#view the final dataframe 
wego |> head(3)

```
Use the date and time criteria to filter the wego dataset, and create a column with 0/1 flag is TSP is used for that specific row 

```{r}
## create a column with the headway calculations (used in the headway conditional fitlter)
wego <- wego |> mutate(pc_hdwy_calc = HDWY_DEV/SCHEDULED_HDWY, .before = HDWY_DEV)

#Create a tsp-on column using case_when to filter on the specific dates and routes
wego_tsp <- wego |> 
  mutate(tsp_on= case_when( 
    #First unconditional day with TSP on 
    between(timestamp, ymd_hms("2025-02-03 12:00:00"), ymd_hms("2025-02-10 12:00:00")) & 
      tsp_eligible == 1 ~ 1,
    #Unconditional day- add additional filter if adherence > 2 (more than 2 minutes late)
    between(timestamp, ymd_hms("2025-02-10 12:00:00"),ymd_hms("2025-04-28 12:00:00")) &
      tsp_eligible == 1 & 
      (ADHERENCE <= -2)~ 1,
    #Second unconditional day
    between(timestamp, ymd_hms("2025-05-05 12:00:00"),ymd_hms("2025-05-12 12:00:00")) &
      tsp_eligible == 1 ~ 1,
    #conditional day- only bus where headway percent is greater than 120%
    timestamp >= ymd_hms("2025-05-12 12:00:00") &
      tsp_eligible == 1 & 
      (pc_hdwy_calc >= 0.2) ~ 1,
    #All other days are considered non-TSP days 
    .default = 0),
    
    #Place the new columns as the first in the dataframe for easier viewing
    .before = CALENDAR_ID
  )

#Count how many tsp-on days were flagged to compare effectiveness of filtering to group members
num_total_tsp_flag <- wego_tsp |> 
  filter(tsp_on == 1) |> 
  nrow()

paste("After filtering, there are ", num_total_tsp_flag, " trips where TSP was used")

```
Add an additonal column for model- was the TSP on, conditional, or unconditional? 
Add a column: 0= TSP off, 1= TSP unconditional, 2= TSP conditional
```{r}
## create a column with the headway calculations (used in the headway conditional fitlter)
wego <- wego |> mutate(pc_hdwy_calc = HDWY_DEV/SCHEDULED_HDWY, .before = HDWY_DEV)

#Create a tsp-on column using case_when to filter on the specific dates and routes
wego_tsp <- wego_tsp |> 
  mutate(tsp_cond_status= case_when( 
    #First unconditional day with TSP on 
    between(timestamp, ymd_hms("2025-02-03 12:00:00"), ymd_hms("2025-02-10 12:00:00")) & 
      tsp_eligible == 1 ~ 1,
    #conditional day- add additional filter if adherence > 2 (more than 2 minutes late)
    between(timestamp, ymd_hms("2025-02-10 12:00:00"),ymd_hms("2025-04-28 12:00:00")) &
      tsp_eligible == 1 ~ 2, #& (ADHERENCE <= -2)
    #Second unconditional day
    between(timestamp, ymd_hms("2025-05-05 12:00:00"),ymd_hms("2025-05-12 12:00:00")) &
      tsp_eligible == 1 ~ 1,
    #conditional day- only bus where headway percent is greater than 120%
    timestamp >= ymd_hms("2025-05-12 12:00:00") &
      tsp_eligible == 1 ~ 2, #& (pc_hdwy_calc >= 0.2)
    #All other days are considered non-TSP days 
    .default = 0),
    
    #Place the new columns as the first in the dataframe for easier viewing
    .before = CALENDAR_ID
  )

wego_tsp
```
**Creating the model for probability calculations**
Does TSP reduce the probability that a bus leaves a stop late?

With a tsp column, we can use that as a variable for a logistical regression. The main model questions would be if the TSP_on being used will predict/impact the bus being late to the stop.

The model will have these structures: 
1) TSP_on having 0 or 1 (was tsp on or off)
2) Was the bus late (ADJUSTED_LATE_COUNT)
3) Additional variables 

First general model: 
target variable = Adjusted_late_count, 
predictor variable = tsp on or off

Result -> probability that the TSP on/off changed bus being late 

Second models: refining comparrison criteria 
* When looking at just the tsp on/off on the same days - portions of the route off vs on 
* looksing at previous history- the same route stops that could have tsp on but (historically/conditionally) they were off 
* accounting for additional factors like time of day, day of week, etc that could impact the model 

```{r}
#quick dataframe reference
wego_tsp|> head(2)

```

Question: Does TSP on/off (conditional/unconditional pooled) predict the amount of time (adherence)
```{r}

## Creating the first general model -> LOGISTICAL
#target = ADHERENCE (how early/late the bus from the previous stop)
#predictor = tsp on/off

tsp_late_model_general <- glm(ADHERENCE~ tsp_on, 
                  data=wego_tsp)

summary(tsp_late_model_general)

```

``` {r}
#create a quick visual 

ggplot(data=wego_tsp,
       aes(x=tsp_on, y=ADHERENCE)
       )+
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title= "Does TSP use impact bus lateness?", 
       x="TSP status", 
       y="Adherence")

```
Adherence does not answer our question- instead we need to look at the adjusted_late_count. So, first let us group by a variable to get total percent of buses that were late each day. 

Grouping 1: What was the percent of late bus arrivals each day, and was this day a TSP on/off day?

```{r}
#Group each date for comparrison 
#First get the info on the percent of buses that were late on that date 
group_late_date <- 
  wego_tsp |> 
  group_by(DATE) |> 
  summarise(pc_late = mean( ADJUSTED_LATE_COUNT))

#Then count the total number of rows that had tsp on 
tsp_count_col<- 
  wego_tsp |> 
  group_by(DATE) |> 
  summarize(tsp_on_count = sum(tsp_on))

#Also count the total number of rows that were conditional or not 
tsp_cond_count_col <- 
  wego_tsp |> 
  group_by(DATE) |> 
  summarize(tsp_cond_count = max(tsp_cond_status)) #pull out the max number from this column

#compile the list together using merge function for graphing 
group_date_tsp_data <- inner_join(
  group_late_date, 
  tsp_count_col, 
  by= 'DATE')
group_date_tsp_data <- inner_join(
  group_date_tsp_data, 
  tsp_cond_count_col, 
  by= 'DATE')

#visualize the resulting dataframe 
group_date_tsp_data
```

```{r}

#create a quick visual graph
ggplot(data=group_date_tsp_data,
       aes(x=tsp_on_count, 
           y=pc_late, 
           color=tsp_cond_count)
       )+
  geom_point() +
  #geom_smooth(method = "lm") +
  labs(title= "Does number of TSP stops improve percent of buses late?",
       x="number TSP on stops",
       y="percent of buses that arrived late")

```
I will remove the TSP=0 days, since it has wide variety. And instead look at the linear relationship between number of TSP stops and percent of late counts for that day 

```{r}

#remove the tsp count = 0 days 
date_late_count_group_dropzero <- group_date_tsp_data |> filter(tsp_on_count > 0) 

#convert the tsp_on_count to a factor in order to make it a discrete variable instead of a continuous variable
date_late_count_group_dropzero <- date_late_count_group_dropzero|> 
  mutate(tsp_cond_count = factor(tsp_cond_count,
                                 levels = c(1,2),
                                 labels = c('unconditional', 'conditional')))

date_late_count_group_dropzero |> head()
```
```{r}
#create a quick visual graph
ggplot(date_late_count_group_dropzero,
       aes(x=tsp_on_count, 
           y=pc_late, 
           color=tsp_cond_count)
       )+
  geom_point() +
  #Add the linear regression line for visual
  geom_smooth(method = "lm", se=FALSE) + 
  labs(title= "Impact of the number of tsp events on percent of daily buses late",
       x="number of events with TSP on",
       y="Percent of late buses", 
       color="Conditional status")  
  #ggsave("Impact of the number of tsp events on percent of daily buses late.png")

```
** Note on analysis: This is not showing what I originally thought**
Essentially it is showing that on conditional days (light blue), the TSP was triggered more because they buses were running late. So the relationship of TSP improving bus timeliness is not explicit, but rather that the conditional TSP settings is working as expected. I will need to restructure my data and question to answer the ACTUAL relationship question of the effectiveness of TSP. Likely by comparing at the route or date month level.
**

Run A linear regression -> 
target= percent of late buses each day 
predictor = number of tsp stops
predictor 2 = conditional or not 

```{r}
date_late_count_group_dropzero |> distinct(tsp_cond_count)

#run a linear regression with the variables
group_date_late_cond_model <- lm(pc_late~ tsp_on_count + tsp_cond_count,
                  data=date_late_count_group_dropzero)

summary(group_date_late_cond_model)
```
Run A LOGISTICAL regression -> group by route??
target= percent of late buses each day 
predictor = TSP on/off
predictor 2 = conditional or not 


```{r}

## Group by Route -> at each route stop, does the tsp on/off effect count of buses late?
daily_late_stop <- wego_tsp |> group_by(TIME_POINT_ABBR, tsp_on, DATE) |> summarise(pc_late = mean( ADJUSTED_LATE_COUNT))


daily_late_stop #|> distinct(tsp_on)

#make TSP on a factor
daily_late_stop <- daily_late_stop|> 
  mutate(tsp_on = factor(tsp_on,
                         levels=c(0,1),
                         labels=(c('off', 'on'))))

daily_late_stop 
ggplot(daily_late_stop,
       aes(x=tsp_on,
           y=pc_late, 
           color=tsp_on)
       )+
geom_point() + 
  geom_smooth(method='lm') +
  facet_grid(cols=vars(TIME_POINT_ABBR))

# avg_daily_late_stop <- daily_late_stop |> group_by(TIME_POINT_ABBR, tsp_on) |> summarize(avg_daily_pc = mean(pc_late))
# 
# avg_daily_late_stop |> head()
# 
# #plot each stop as on/off and the pc late
# ggplot(avg_daily_late_stop,
#        aes(x=TIME_POINT_ABBR,
#            y=avg_daily_pc,
#            color=tsp_on)
#        )+
#   geom_bar() #+
#   #Add the linear regression line for visual
#   #geom_smooth(method = "lm", se=FALSE) # +
#   # labs(title= "Impact of the number of tsp events on percent of daily buses late",
#   #      x="number of events with TSP on",
#   #      y="Percent of late buses",
#   #      color="Conditional status")
#   # #ggsave("Impact of the number of tsp events on percent of daily buses late.png")

```
----
What if we just look at the conditional/unconditional differences? Does that improve the adherence predictibility?

```{r}
# wego_tsp |> head()
```

```{r}
## Aditional model-> create conditional tsp column as a categorical variable in the model 

# # #optional setup= need to make it a factor (from titanic warmup exercise?)
# 
# wego_tsp$tsp_cond_status <- factor(
#   wego_tsp$tsp_cond_status, 
#   levels = c("off", "conditional", "unconditional"), 
#   labels = c(0, 1, 2))
# 
# 
# # 
# tsp_late_model_conditional <- glm(ADHERENCE~ tsp_on + tsp_cond_status,
#                   data=wego_tsp, 
#                   family= "binomial")
# 
# summary(tsp_late_model_conditional)
# 
# # 
# # #create a quick visual
# # 
# # ggplot(data=wego_tsp,
# #        aes(x=tsp_cond_status, y=ADHERENCE)
# #        )+
# #   geom_point() +
# #   #geom_smooth(method = "lm") +
# #   labs(title= "Does TSP use impact bus lateness?",
# #        x="TSP condiitonal status",
# #        y="Adherence")
# 
# 
# wego_tsp
```
